Fast Sorting on Modern Computers

### 2005-05-28 paper/docs/method.tex
cleaned up the method

\label{method}
%TODO rewrite this
This section discusses the method and framework used to complete this project.
From the point of view of a Final Year Project, this is a description of what I
did all year, what I spent my time on, what went right and wrong during the
project, and what I learned. The first section discusses the steps involved in
creating the project, programming the sorts and getting results. The second
section discusses tools that I used or created in the course of the project.

\section{Method}
i
%TODO remove this
The following is the initial project description:
\begin{quotation}
Sorting is one of the most important and studied problems in computer science.
Many good algorithms exist which offer various trade-offs in efficiency,
simplicity, and memory use.  One weakness of classical sorting algorithms is
that they are not designed for modern computers.

Cache memories, pipelines, and branch prediction have greatly changed the
relative strengths of different algorithms. For example, comparison based
sorting algorithms (such as Quicksort) can be inefficient on modern machines,
because they involve difficult to predict branches.

The goal of this project is to design, implement and experiment with sorting
algorithms designed to make the best use of modern architectural features.
Several cache-conscious sorting algorithms have been proposed, and the first
step would be to implement these algorithms and compare their performance with
traditional algorithms.  The next step is to consider modifications to the
algorithms to make better use of the processor pipeline and multiple execution
units and reduce branch mispredictions. The project will also involve a detailed
study of the behaviour of the algorithms using a processor simulator or
performance counters on a real machine.

To do this project, you should have a reasonable knowledge of computer
architecture and traditional sorting algorithms. You should also be enthusiastic
about algorithm design, and making things go faster.  All programming is in C or
C++.
\end{quotation}
After discussion with my supervisor, the following project goals were agreed upon:
\begin{itemize}
\item Implement heapsort, mergesort, quicksort, radixsort, and the four
elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
\item Simulate these sorts using the SimpleScalar architecture simulator, using a
variety of branch predictors and cache configurations.
\item Implement cache-conscious sorts from \cite{LaMarca96} and
\cite{LaMarca99}.
\item Investigate and analyse the increase in speed, reduction in cache misses, 
and change in branch prediction accuracy as a result of these changes.
\end{itemize}

Several steps were necessary to prepare for the project. The first of these was
to compile SimpleScalar. A guide to repeating this is included in Section
\ref{SimpleScalar Install}. A framework was required by which I could test sorts
as they were developed. This is discussed in Section \ref{Main}. It was also necessary
to fully understand the technologies to be used in the project. Several texts
were used for this: cache and branch predictors are both in
\cite{HennessyPatterson90}; branch predictors in
\cite{Uht97}, as well as in \cite{McFarling93}
and \cite{Smith81}; sorts are discussed in \cite{Sedgewick02} and
\cite{Knuth98}; cache-conscious sorts are discussed \cite{LaMarca96},
\cite{LaMarca99} and \cite{LaMarcaHeap96}.

There were several steps involved in writing each iteration of the algorithms.
The elementary sorts and radixsort had just one of these iterations; the other sorts had more
than one\footnote{Quicksort, for example, had four: base quicksort, tuned
quicksort, multi-quicksort and sequential multi-quicksort.}. A flowchart of these
steps appears in Figure \vref{Flowchart}.

\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

The first step was to write the algorithm. \cite{Sedgewick02} was the primary
source of these, though in several cases LaMarca specified a particular version
which he used. When a step involved in an algorithm wasn't clear, the paper
describing the algorithm was used to clarify the meaning.

@@

While it was not known in advance the desired instruction count of an algorithm,
a certain range was implied by LaMarca's descriptions. The relationship between
one iteration of an algorithm, or between base versions of several algorithms was
known ahead of time. One configuration of the \cc{simple} script (see
Section \ref{Simple Script}) ran quick tests using the \cc{sim-fast} simulator,
giving instruction counts which could be compared to other versions.

@@

The installation of SimpleScalar can be difficult. A step by step description
for installing it on an x86 linux machine is in Section \ref{SimpleScalar
Install}.

@@

The output of SimpleScalar is human readable (see Section \ref{Sample
Simulation} for an example) but is not conducive to collecting large amounts of
data. To that end, a script was written to run each required set of simulations. The
script was written in perl, and recorded data from the simulations into a single
file for later retrieval. The source code to the script, \cc{simple}, is in
section \ref{Simple Script}.  A sample list of commands executed by \cc{simple} are
in section \ref{Simple List}. The data file produced was suitable to reading
with a \cc{diff} program.

@@

There were several steps involved in using gnuplot for the data. Firstly, the
data was not in the correct format for gnuplot, and had to be converted. It also
needed to be offset by the cost of filling the array with random variables, and
scaled to so that the y-axis could be in the form of \n{plotted data per key},
both to be more readable, and to be consistent with LaMarca's graphs. The source
code to the script to do this is in Figure \ref{Convert Script}. Once the data
was converted, a script to run gnuplot (see Section \ref{Sample Gnuplot Script}
for an example) was required. The number of graphs required made this an
impossible manual task, and so another script had to be written.  Some graphs
still needed to be tweaked by hand, due to difficulties in the script, but the
task was much shortened by this, and the tweaking was mostly accomplished using
macros in the development platform.

@@

\subsection{Incidental Tools}
The algorithms in this project were written in C using \cc{vim}, a powerful text
editor. \cc{gcc} was used to compile the framework, when not using SimpleScalar.
The report was written using \LaTeX, and diagrams were written is
\cc{postscript}, or drawn using \cc{OpenOffice.org}, an open-source office
suite. Many files had to be edited in the same way, and this was accomplished
either using \cc{vim} macros, or by writing a short \cc{perl} script.

@@

The \cc{C} code written for this project was approximately 8000 lines long.
Appendix \ref{Sort Listings} lists many of the key parts of this code.

@@

Each of the three $O(NlogN)$ sorts was treated in this fashion. The mergesorts
accounted for over 3000 lines of the 8000 lines of code in this project. Key
portions of this code are shown in Section \ref{Sort Listings}.

>>>>

\label{method}
This section discusses the method and framework used to complete this research.
The first section discusses the steps involved in creating a testing framework,
programming the sorts and getting results. The second section discusses tools
that I used or created to do this.

\section{Method}

The following were the objectives of the research:
\begin{itemize}
\item Implement heapsort, mergesort, quicksort, radixsort, shellsort and the four
elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
\item Simulate these sorts using the SimpleScalar architecture simulator, using a
variety of branch predictors and cache configurations.
\item Implement cache-conscious sorts from \cite{LaMarca96} and
\cite{LaMarca99}.
\item Investigate and analyse the increase in speed, reduction in cache misses, 
and change in branch prediction accuracy as a result of these changes.
\item Determine if any changes can be made to improved the sorts' performance.
\end{itemize}

% TODO maybe remoe this?
Several steps were necessary to prepare for the research.  A framework was
required by which sorts could be tested as they were being developed. This is
discussed in Section \ref{Main}. It was also necessary to fully understand the
technologies to be used in the project. Several texts were used for this: cache
and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.

There were many steps involved in writing each iteration of the algorithms.
Insertion and selection sort had just one of these iterations; the other
sorts had more than one\footnote{Quicksort, for example, had four: base
quicksort, tuned quicksort, multi-quicksort and sequential multi-quicksort.}. A
flowchart of these steps appears in Figure \vref{Flowchart}.

%TODO this is going to have to be updated
\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

The first step in each case was to write the algorithm. \cite{Sedgewick02} was the primary
source of these, though in several cases LaMarca specified a particular version
which he used. When a step involved in an algorithm wasn't clear, the paper
describing the algorithm was used to clarify the meaning.

@@

While it was not known in advance the desired instruction count of an algorithm,
a certain range was implied by LaMarca's descriptions. The relationship between
one iteration of an algorithm, or between base versions of several algorithms
was known ahead of time. One configuration of the \cc{simple} script ran quick
tests using the \cc{sim-fast} simulator, giving instruction counts which could
be compared to other versions.

@@

@@

The output of SimpleScalar is human readable, but is not conducive to collecting
large amounts of data. To that end, a script was written to run each required
set of simulations. It recorded data from the simulations into a single file for
later retrieval. This file was suitable to reading with a \cc{diff} program.

@@

There were several steps involved in using gnuplot for the data. Firstly, the
data was not in the correct format for gnuplot, and had to be converted. It also
needed to be offset by the cost of filling the array with random variables, and
scaled to so that the y-axis could be in the form of \n{plotted data per key},
both to be more readable, and to be consistent with LaMarca's graphs. Once the data
was converted, it was necessary to generate a gnuplot script for each graph.
The number of graphs required made this an
impossible manual task, and so another script had to be written.  Some graphs
still needed to be tweaked by hand, due to difficulties in the script, but the
task was much shortened by this, and the tweaking was mostly accomplished using
macros in the development platform.

@@

@@

@@

%TODO another sentence here
Each of the four $O(NlogN)$ sorts and radixsort, was treated in this fashion.
